Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-03-06 article.tex
Use bibtex instead of what quarto generated

Several tools already exist for defining and simulating \acp{CRN}.
BioSimulators.org (Shaikh et al. 2022), a registry of simulation tools,
list at least 15 Python software including COPASI (Hoops et al. 2006),
Tellurium (Choi et al. 2018) and PySB (Lopez et al. 2013). COPASI is a
standalone software with a \ac{GUI} for defining and simulating models. It's
widely used for its user-friendly interface and comprehensive features.
There are also python bindings and interfaces for COPASI to allow
advanced scripting. Tellurium is a Python-based modeling environment,
however, it uses a C++ library called libRoadRunner in the backend to
simulate models. PySB is a Python library that created a \ac{DSL} using
standard Python to define models, which are then compiled to \acp{ODE} using a
Perl library called BioNetGen (L. A. Harris et al. 2016).

@@

Another challenge is the way models are written. Many tools use a \ac{DSL} for coding, and support the
\ac{SBML} as an exchange format (Hucka et al., n.d.), as
direct \ac{SBML} coding is impractical. A \ac{DSL} allows reuse of code in
different programming environments, but it will not be recognized by
default in \acp{IDE}. Therefore \acp{DSL}
cannot provide the development experience such as syntax highlighting,
code completion, refactoring, and static analysis, unless such support
is specifically developed. Tellurium uses a
\ac{DSL} called Antimony (Smith et al. 2009), which can be translated to
\ac{SBML}. An extension for Visual Studio Code was developed, but its
maintenance could be a demanding task for the systems biology community.
PySB, using Python's dynamic nature, created a \ac{DSL} within Python. By
default, it uses global state to create species and parameters, without
assigning them to Python variables or adding them explicitly to the
model, but this approach is not fully compatible with \acp{IDE}, affecting
the development experience.

@@

Modular code architecture makes code reusable, extensible, and easier to
maintain. Therefore, we split the code to define and simulate reaction
systems into three Python packages: symbolite, to create symbolic
expressions; poincaré, to define dynamical systems; and simbio, to
define reaction systems and interface with systems biology standards
such as \ac{SBML}. These are pure Python packages with standard dependencies
from the PyData scientific stack such as NumPy (C. R. Harris et al.
2020) and pandas (McKinney 2010). They are published in the \acf{PyPI}
, where links to the source code and documentation hosted
in GitHub can be found, and can be easily installed with
\texttt{pip\ install\ \textless{}package\_name\textgreater{}}.

Symbolite is a lightweight symbolics package to create algebraic
mathematical expressions. Symbolite expressions can be inspected and
compiled to various backends. Currently, we have implementations for
NumPy (C. R. Harris et al. 2020); Numba (Lam, Pitrou, and Seibert 2015),
a \ac{JIT} compiler to LLVM; SymPy (Meurer et al. 2017), a
library for symbolic mathematics; and JAX (Bradbury et al. 2018), a
library that support automatic differentiation and compilation to \acp{GPU}
and \acp{TPU}. Symbolite is designed to facilitate the easy integration of
new backends.

@@

For this last point, \acp{IDE} that support \texttt{dataclass\_transform} (De
Bonte and Traut 2021) can provide a tooltip with the expected signature
(Figure~\ref{fig-ide}). This requires the use of type
annotations, which play a more significant role in static type checking
as they can help to identify errors before running the code. For
instance, to parameterize the initial conditions of variables we have to
use a \texttt{Constant}. If we try to use a \texttt{Parameter}, which
could be a time-dependent expression, it is flagged as a type error
(Figure~\ref{fig-ide}).

@@

Several commonly used reactions are predefined as \texttt{MassAction}
subclasses, such as \texttt{MichaelisMenten}
(\(S + E \leftrightarrow ES \rightarrow P + E\)) and its approximate
form without the intermediate species \(ES\), and it is also simple to
implement your own as subclasses of \texttt{RateLaw} or
\texttt{MassAction}. Additionally, SimBio supports importing from and
exporting to \ac{SBML}, and downloading them directly from BioModels
(Malik-Sheriff et al. 2020) (Figure~\ref{fig-simbio-io}).

@@

To evaluate SimBio's reproducibility, we analyzed curated \ac{SBML} models
from BioModels (Malik-Sheriff et al. 2020). Among the first 100 curated
models from BioModels we selected 60 that did not contains events, as
SimBio doesn't support events yet. We simulated the selected models with
COPASI and used the simulated results as ground truth, and demonstrated
that SimBio reproduces the results with minor numerical differences
depending on the solver tolerance.

@@

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-jax2018github}{}}%
Bradbury, James, Roy Frostig, Peter Hawkins, Matthew James Johnson,
Chris Leary, Dougal Maclaurin, George Necula, et al. 2018. {``{JAX}:
Composable Transformations of {Python}+{NumPy} Programs.''}

\leavevmode\vadjust pre{\hypertarget{ref-choiTelluriumExtensiblePythonbased2018}{}}%
Choi, Kiri, J Kyle Medley, Matthias König, Kaylene Stocking, Lucian
Smith, Stanley Gu, and Herbert M Sauro. 2018. {``Tellurium: {An}
Extensible Python-Based Modeling Environment for Systems and Synthetic
Biology.''} \emph{Bio Systems} 171 (September): 74--79.
\url{https://doi.org/10.1016/j.biosystems.2018.07.006}.

\leavevmode\vadjust pre{\hypertarget{ref-debontePEP681Data2021}{}}%
De Bonte, Erik, and Eric Traut. 2021. {``{PEP} 681 {\textendash} {Data
Class Transforms}.''} \emph{Python Enhancement Proposals}.
https://peps.python.org/pep-0681/.

\leavevmode\vadjust pre{\hypertarget{ref-harrisArrayProgrammingNumPy2020}{}}%
Harris, Charles R., K. Jarrod Millman, Stéfan J. van der Walt, Ralf
Gommers, Pauli Virtanen, David Cournapeau, Eric Wieser, et al. 2020.
{``Array Programming with {NumPy}.''} \emph{Nature} 585 (7825): 357--62.
\url{https://doi.org/10.1038/s41586-020-2649-2}.

\leavevmode\vadjust pre{\hypertarget{ref-harrisBioNetGenAdvancesRulebased2016}{}}%
Harris, Leonard A, Justin S Hogg, José-Juan Tapia, John A P Sekar,
Sanjana Gupta, Ilya Korsunsky, Arshi Arora, Dipak Barua, Robert P
Sheehan, and James R Faeder. 2016. {``{BioNetGen} 2.2: Advances in
Rule-Based Modeling.''} \emph{Bioinformatics (Oxford, England)} 32 (21):
3366--68. \url{https://doi.org/10.1093/bioinformatics/btw469}.

\leavevmode\vadjust pre{\hypertarget{ref-hoopsCOPASICOmplexPAthway2006}{}}%
Hoops, Stefan, Sven Sahle, Ralph Gauges, Christine Lee, Jürgen Pahle,
Natalia Simus, Mudita Singhal, Liang Xu, Pedro Mendes, and Ursula
Kummer. 2006. {``{COPASI}{\textemdash}a {COmplex PAthway SImulator}.''}
\emph{Bioinformatics} 22 (24): 3067--74.
\url{https://doi.org/10.1093/bioinformatics/btl485}.

\leavevmode\vadjust pre{\hypertarget{ref-huckaSBMLL3V2}{}}%
Hucka, Michael, Frank Bergmann, Stefan Hoops, Sarah M Keating, Nicolas
Le Novère, Chris J Myers, Brett G Olivier, et al. n.d. {``{SBML
L3V2}.''}

\leavevmode\vadjust pre{\hypertarget{ref-lamNumbaLLVMbasedPython2015}{}}%
Lam, Siu Kwan, Antoine Pitrou, and Stanley Seibert. 2015. {``Numba: A
{LLVM-based Python JIT} Compiler.''} In \emph{Proceedings of the {Second
Workshop} on the {LLVM Compiler Infrastructure} in {HPC}}, 1--6. {LLVM}
'15. {New York, NY, USA}: {Association for Computing Machinery}.
\url{https://doi.org/10.1145/2833157.2833162}.

\leavevmode\vadjust pre{\hypertarget{ref-lopezProgrammingBiologicalModels2013}{}}%
Lopez, Carlos F, Jeremy L Muhlich, John A Bachman, and Peter K Sorger.
2013. {``Programming Biological Models in {Python} Using {PySB}.''}
\emph{Molecular Systems Biology} 9 (January): 646.
\url{https://doi.org/10.1038/msb.2013.1}.

\leavevmode\vadjust pre{\hypertarget{ref-malik-sheriffBioModels15Years2020}{}}%
Malik-Sheriff, Rahuman S, Mihai Glont, Tung V N Nguyen, Krishna Tiwari,
Matthew G Roberts, Ashley Xavier, Manh T Vu, et al. 2020.
{``{BioModels}{\textemdash}15 Years of Sharing Computational Models in
Life Science.''} \emph{Nucleic Acids Research} 48 (D1): D407--15.
\url{https://doi.org/10.1093/nar/gkz1055}.

\leavevmode\vadjust pre{\hypertarget{ref-mckinneyDataStructuresStatistical2010}{}}%
McKinney, Wes. 2010. {``Data {Structures} for {Statistical Computing} in
{Python}.''} In \emph{Python in {Science Conference}}, 56--61. {Austin,
Texas}. \url{https://doi.org/10.25080/Majora-92bf1922-00a}.

\leavevmode\vadjust pre{\hypertarget{ref-meurerSymPySymbolicComputing2017}{}}%
Meurer, Aaron, Christopher P. Smith, Mateusz Paprocki, Ondřej Čertík,
Sergey B. Kirpichev, Matthew Rocklin, AmiT Kumar, et al. 2017.
{``{SymPy}: Symbolic Computing in {Python}.''} \emph{PeerJ Computer
Science} 3 (January): e103. \url{https://doi.org/10.7717/peerj-cs.103}.

\leavevmode\vadjust pre{\hypertarget{ref-shaikhBioSimulatorsCentralRegistry2022}{}}%
Shaikh, Bilal, Lucian P Smith, Dan Vasilescu, Gnaneswara Marupilla,
Michael Wilson, Eran Agmon, Henry Agnew, et al. 2022.
{``{BioSimulators}: A Central Registry of Simulation Engines and
Services for Recommending Specific Tools.''} \emph{Nucleic Acids
Research} 50 (W1): W108--14. \url{https://doi.org/10.1093/nar/gkac331}.

\leavevmode\vadjust pre{\hypertarget{ref-smithAntimonyModularModel2009}{}}%
Smith, Lucian P., Frank T. Bergmann, Deepak Chandran, and Herbert M.
Sauro. 2009. {``Antimony: A Modular Model Definition Language.''}
\emph{Bioinformatics} 25 (18): 2452--54.
\url{https://doi.org/10.1093/bioinformatics/btp401}.

\end{CSLReferences}

>>>>

Several tools already exist for defining and simulating \acp{CRN}.
BioSimulators.org \cite{shaikhBioSimulatorsCentralRegistry2022}, a registry of simulation tools,
list at least 15 Python software including COPASI \cite{hoopsCOPASICOmplexPAthway2006},
Tellurium \cite{choiTelluriumExtensiblePythonbased2018} and PySB \cite{lopezProgrammingBiologicalModels2013}. COPASI is a
standalone software with a \ac{GUI} for defining and simulating models. It's
widely used for its user-friendly interface and comprehensive features.
There are also python bindings and interfaces for COPASI to allow
advanced scripting. Tellurium is a Python-based modeling environment,
however, it uses a C++ library called libRoadRunner in the backend to
simulate models. PySB is a Python library that created a \ac{DSL} using
standard Python to define models, which are then compiled to \acp{ODE} using a
Perl library called BioNetGen \cite{harrisBioNetGenAdvancesRulebased2016}.

@@

Another challenge is the way models are written. Many tools use a \ac{DSL} for coding, and support the
\ac{SBML} as an exchange format \cite{huckaSBMLL3V2}, as
direct \ac{SBML} coding is impractical. A \ac{DSL} allows reuse of code in
different programming environments, but it will not be recognized by
default in \acp{IDE}. Therefore \acp{DSL}
cannot provide the development experience such as syntax highlighting,
code completion, refactoring, and static analysis, unless such support
is specifically developed. Tellurium uses a
\ac{DSL} called Antimony \cite{smithAntimonyModularModel2009}, which can be translated to
\ac{SBML}. An extension for Visual Studio Code was developed, but its
maintenance could be a demanding task for the systems biology community.
PySB, using Python's dynamic nature, created a \ac{DSL} within Python. By
default, it uses global state to create species and parameters, without
assigning them to Python variables or adding them explicitly to the
model, but this approach is not fully compatible with \acp{IDE}, affecting
the development experience.

@@

Modular code architecture makes code reusable, extensible, and easier to
maintain. Therefore, we split the code to define and simulate reaction
systems into three Python packages: symbolite, to create symbolic
expressions; poincaré, to define dynamical systems; and simbio, to
define reaction systems and interface with systems biology standards
such as \ac{SBML}. These are pure Python packages with standard dependencies
from the PyData scientific stack such as NumPy \cite{harrisArrayProgrammingNumPy2020}
and pandas \cite{mckinneyDataStructuresStatistical2010}. They are published in the \acf{PyPI}
, where links to the source code and documentation hosted
in GitHub can be found, and can be easily installed with
\texttt{pip\ install\ \textless{}package\_name\textgreater{}}.

Symbolite is a lightweight symbolics package to create algebraic
mathematical expressions. Symbolite expressions can be inspected and
compiled to various backends. Currently, we have implementations for
NumPy \cite{harrisArrayProgrammingNumPy2020}; Numba \cite{lamNumbaLLVMbasedPython2015},
a \ac{JIT} compiler to LLVM; SymPy \cite{meurerSymPySymbolicComputing2017}, a
library for symbolic mathematics; and JAX \cite{jax2018github}, a
library that support automatic differentiation and compilation to \acp{GPU}
and \acp{TPU}. Symbolite is designed to facilitate the easy integration of
new backends.

@@

For this last point, \acp{IDE} that support \texttt{dataclass\_transform} \cite{debontePEP681Data2021}
can provide a tooltip with the expected signature
(Figure~\ref{fig-ide}). This requires the use of type
annotations, which play a more significant role in static type checking
as they can help to identify errors before running the code. For
instance, to parameterize the initial conditions of variables we have to
use a \texttt{Constant}. If we try to use a \texttt{Parameter}, which
could be a time-dependent expression, it is flagged as a type error
(Figure~\ref{fig-ide}).

@@

Several commonly used reactions are predefined as \texttt{MassAction}
subclasses, such as \texttt{MichaelisMenten}
(\(S + E \leftrightarrow ES \rightarrow P + E\)) and its approximate
form without the intermediate species \(ES\), and it is also simple to
implement your own as subclasses of \texttt{RateLaw} or
\texttt{MassAction}. Additionally, SimBio supports importing from and
exporting to \ac{SBML}, and downloading them directly from BioModels
\cite{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).

@@

To evaluate SimBio's reproducibility, we analyzed curated \ac{SBML} models
from BioModels \cite{malik-sheriffBioModels15Years2020}. Among the first 100 curated
models from BioModels we selected 60 that did not contains events, as
SimBio doesn't support events yet. We simulated the selected models with
COPASI and used the simulated results as ground truth, and demonstrated
that SimBio reproduces the results with minor numerical differences
depending on the solver tolerance.

@@

\bibliographystyle{plain}
\bibliography{bibliography.bib}
